Staged Abstract Interpreters

### 2019-08-15 paper/50_stageabs.tex
revision

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of their
confluence -- a staged abstract interpreter.
Unsurprisingly, the staged abstract interpreter in this section has the same
abstract semantics with the unstaged version from Section~\ref{unstaged_abs}.
The approach from unstaged to staged abstract interpreter is modular,
principled, and does not sacrifice soundness or precision. The designer of the
analyzer has no need to rewrite the analysis. We first present the staged
lattices and staged monads, and then the staged version of primitive operations,
especially @close@, @ap_clo@ and @fix@, at last, we discuss several optimizations
and our design.

@@

\begin{lstlisting}
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: Rep[Set[T]] = Set[T]()
    lazy val ⊤: Set[T] = error("No representation for ⊤")
    def ⊑(l1: Rep[Set[T]], l2: Rep[Set[T]]): Rep[Boolean] = l1 subsetOf  l2
    def ⊔(l1: Rep[Set[T]], l2: Rep[Set[T]]): Rep[Set[T]]  = l1 union     l2
    def ⊓(l1: Rep[Set[T]], l2: Rep[Set[T]]): Rep[Set[T]]  = l1 intersect l2
  }
\end{lstlisting}

As we can see, the @RepSetLattice@ is an instance of @RepLattice@, which is also
an abstraction of the underlying data structure. The lattice operations
eventually are delegated to @subsetOf@, @union@ and @intersect@ on
@Rep[Set[T]]@. We even don't have to change the implementation code, but just
the types – from @Set[T]@ to @Rep[Set[T]]@. LMS library provides an implicit
conversion lifting to next-stage values. In the code generation part, these
operations emit their corresponding next-stage code. Again, other lattice
structures such as products and maps are implemented in a similar way.

@@

Now we implement the staged abstract semantics, where @R@ is instantiated as
@Rep@, and the abstract components are shared from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as in the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For the readability, we squeeze the three
inner transformers into a single monad @RepSetReaderStateM[A, B, C]@, where @A@
and @B@ are both @Cache@ when used in @AnsM@. Now the result type is a pair of
two staged value: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type RepSetReaderStateM[A, B, C] = RepSetT[RepReaderT[RepStateT[RepIdM, A, ?], B, ?], C]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

Since our monad stack is uniformly using staged data, so the staged @SetT@ now
stores a staged value of type @Rep[Set[A]]@, inside of another staged monad @M@.

\begin{lstlisting}
  case class SetT[M[_]: RepMonad, A](run: M[Set[A]]) {
    def flatMap[B: Manifest](f: Rep[A] => SetT[M, B]): SetT[M, B] = ...
  }
\end{lstlisting}

\paragraph{Primitive Operations} We keep our road map that focuses on functions
and applications. The @close@ method now is a rough mixing of the staged
concrete and unstaged abstract version: we have a current-stage function @f@,
that takes four next-stage values, @in@ and @out@ additionally; inside of @f@,
we collapse the @Ans@ monad to values of type @Result@ by providing the desired
arguments, i.e., the new environment, new store, and caches. The collapse
happens at current-stage, so the call of @ev@ is unfolded after staging. At
last, we generate a singleton set containing the compiled closure
@emit_compiled_clo(f)@, represented by an IR node in LMS.

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the set of closures into the monad stack, and generate a next-stage value
using @emit_ap_clo@, which represents the result of application from future.
Finally, we put the future @out@ cache, store, and values back into the
current-stage monads. The @emit_ap_clo@ method takes the target closure @clo@,
argument @rand@, store, and caches.

\begin{lstlisting}
  def lift_nd[T](vs: Rep[Set[T]]): AnsM[T]
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[(Value, Store)], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store;     clo <- lift_nd[AbsValue](rator)
    in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(Set[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

\paragraph{Staged Caching and Fixpoint Iteration}

Our fixed-point iteration again relies on the two caches @in@ and @out@, which are
both staged values now. Therefore, the query of whether the @out@ cache contains
the current configuration produces a next-stage Boolean value, i.e.,
@Rep[Boolean]@. Consequently, the branching cannot be determined statically --
we need to generate code for @if@. Figure \ref{fig:staged_coind_cache} shows the
staged version of @fix@.

\vspace{-1em}
\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(Set[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg), out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}
\vspace{-1em}

The @res@ variable is a next-stage result, which consists of a next-stage @if@
expression. The true branch simply returns a pair of queried result form the
@out@ cache and the @out@ cache. The else branch constructs a monad @m@ first,
which evaluates @e@ under the new @out@ cache. After which, we collapse the
monad @m@ with desired arguments. Finally, we obtain a placeholder @res@ since
it is a next-stage value, and put the content of @res@ back into the monad
stack.

\paragraph{Code Generation} The IR nodes for compiled closures and
applications are defined as follows:

\begin{lstlisting}
  case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value, Store)], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
  case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                     in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

The code generator for these two IR nodes is similar to staged concrete
interpreter, which generates a next-stage @CompiledClo@ object and a next-stage
function call to @f@, respectively.

@@

Our staging schema works by unfolding the interpreter over the input AST. In
practice, it would suffer from code explosion when analyzing (specializing) large
programs, which increases the compile time.
If we generate next stage programs running on JVM, such large generated program
would also cause runtime GC overhead. In this section, we present several
optimizations that largely mitigate these issues. For all that, implementing
these optimizations do not need to change the generic interpreter.

@@

\paragraph{Selective Caching} It is observed that the two-fold co-inductive
caching is used for every recursive call of our abstract interpreter. But this
is not necessary and even very expensive when generating code for atomic
expressions such as literals or variables, as they always terminate. Borrowing
the idea from the partition of A-Normal Form \cite{Flanagan:1993:ECC:155090.155113},
we may use a selective caching algorithm:

@@

\todo{1. construct generic monadic interpreter that abstract over binding-time,
  primtive operations, return type.
  2. implement unstaged abstract interpreter with monads, ie. using approriate
  monad stack
  3. if we want to generate code, we just need to change the monad stack.
  the data wrapped are staged.
  we show readerT monad, stateT monad, fused monad with nondeterminism. }

@@

\todo{may be show an small example?}

@@

As the central concern of static analysis, the soundness is important for
every user who would like to use it. Here we briefly summarize that how our
approach preserve the corretness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here are based on empirical evidence, a rigorous proof of soundness
perservation of the staged abstract interpreters is still an open question.

>>>>

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of their
confluence -- a staged abstract interpreter.
Unsurprisingly, the staged abstract interpreter in this section has the same
abstract semantics with the unstaged version from Section~\ref{unstaged_abs}.
The approach from unstaged to staged abstract interpreter is modular,
and does not sacrifice soundness or precision. The designer of the
analyzer therefore has no need to rewrite the analysis. We first present the
staged lattices and staged monads, and then the staged version of primitive
operations, especially @close@, @ap_clo@ and @fix@. At last, we discuss several
optimizations and summarize our approach.

@@

\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is obtained by specializing @Lattice@ with @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
eventually are delegated to operations on a staged set data type, including
@subsetOf@, @union@ and @intersect@. We even do not have to change the
implementation code, but just the types -- from @Set[T]@ to @Rep[Set[T]]@, shown
in light gray code. The LMS library provides an implicit conversion that lifts a
current-stage set value to next-stage values. In the code generation part, these
operations emit their corresponding next-stage code. Again, other lattice
structures such as products and maps are implemented in a similar way.

@@

Now we can implement the staged abstract semantics, where @R@ is instantiated as
@Rep@, and the abstract components are reused from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For readability, we squeeze the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for reader effect and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with @Cache@. Now the
result type is a pair of two staged value: @Rep[Set[(Value, Store)]]@ and
@Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

\iffalse
Since our monad stack is uniformly using staged data, the staged @SetT@ now
stores a staged value of type @Rep[Set[A]]@, inside of another staged monad @M@.

\begin{lstlisting}
  case class SetT[M[_]: RepMonad, A](run: M[Set[A]]) {
    def flatMap[B: Manifest](f: Rep[A] => SetT[M, B]): SetT[M, B] = ...
  }
\end{lstlisting}
\fi

\paragraph{Primitive Operations} When deriving the staged concrete interpreter
from the unstaged counterpart, we notice that recursive calls of @ev@ on the
lambda body are shifted from @ap_clo@ to @close@. In other word, we need to
eagerly specialize the interpreter and generate code every time when we see a
$\lambda$ term, instead of lazily when we actually need to apply a function.
With no surprise, it is similar in the staged abstract interpreter,
additionally, we need to handle nondeterminism. So we keep our track and focus
on discussing closure representations and function applications for the staged
version.

The @close@ method now is a mixing of the \textit{staged} concrete version and
unstaged \textit{abstract} version: we build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the @Ans@ monad to
grounded values of type @Result@ by providing the desired arguments, i.e., the
new environment, new store, and caches. The collapsing of monad happens at
current-stage, so the call of @ev@ is unfolded through staging. Finally, we
generate a singleton set containing the compiled next-stage closure
@emit_compiled_clo(f)@, represented by an IR node in LMS.

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure @clo@ in the set, by using @emit_ap_clo@. The
method @emit_ap_clo@ generates the current-stage representation of the
future-stage result of application.
Finally, we put the future @out@ cache, store, and values back into
the current-stage monads.

\begin{lstlisting}
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](rator)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2)
    vs  <- lift_nd[ValSt](res._1)
    _   <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

% Monad is the bridge that connects future-stage values

\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix_cache(e: Expr): Ans = for {
    ρ   <- ask_env;  σ <- get_store;  in <- ask_in_cache;  out <- get_out_cache
    cfg <- lift_nd[Config](Set((unit(e), ρ, σ)))
    res <- lift_nd[(Set[ValSt], Cache)](Set(
      // a next-stage value of type Rep[(Set[ValSt], Cache)] generated by if
      if (out.contains(cfg)) (repMapToMapOps(out).apply(cfg).toList, out)
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- eval(fix_cache)(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ)) 
             } yield v
             m(ρ)(σ)(in)(out) }))
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1);  _ <- put_store(vs._2)
  } yield vs._1 
\end{lstlisting}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}

\paragraph{Staged Caching and Fixpoint Iteration} Our fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @out@ cache contains the current
configuration will produce a next-stage Boolean value, i.e., of type
@Rep[Boolean]@. Consequently, the branching condition cannot be determined
statically -- we need to generate code for @if@. Figure
\ref{fig:staged_coind_cache} shows the staged version of @fix@. The
monadic-binded variable @res@ is a next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of queried result form
the @out@ cache and the @out@ cache itself. The else branch constructs a monadic
value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@ cache.
After which, we use the similar technique that collapses the monad @m@ to
grounded values by providing its desired environment and etc. Finally, we have a
current-stage representation of the future-stage values @res@, and put the
content of @res@ back into the monad stack.

\subsection{A Little Bit Code Generation, Again}
The code generation for compiled closures and function applications are similar
to the staged concrete interpreter. We have two the IR node as case classes, they
take additional caches as arguments. We also elide how to generate next-stage Scala
code for them.

\begin{lstlisting}
case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                          => Rep[(Set[ValSt], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                    in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

@@

Our staging schema works by unfolding the interpreter over the input AST. In
practice, however, it would suffer from code explosion when analyzing
(specializing) large programs, which increases the compile time. If we generate
next stage programs running on JVM, such large generated program would also
cause runtime GC overhead. In this section, we present several optimizations
that largely mitigate these issues. For all that, implementing these
optimizations do not need to change the generic interpreter.

@@

\paragraph{Selective Caching} It is observed that the two-fold co-inductive
caching is used for every recursive call of our abstract interpreter. But this
is not necessary and even very expensive when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
\cite{Flanagan:1993:ECC:155090.155113}, we may use a selective caching
algorithm that does not generate caching for atomic expressions:

@@

We summarize our approach to staging an abstract interpreter as follows:

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding-time,
    value domains and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style, therefore the semantics can be encapsulated
    into monads.
  \item By using an approriate moand stack, we can implement a modular, unstaged
    abstract interpreter. This step has been explored in the previous literature.
  \item Replacing the monad stack to staged monad stack, such monads operate on
    staged data types, i.e., next-stage values. We also need to reimplement several
    primitive operations that on staged values, modularly.
\end{itemize}

Monadic interperters are known to be able to decouple the interpretation process
and the underlying semantics. The key insight in this paper is, by making the
monadic interpreter binding-time polymorphic, it can be extended for generating
efficient code in a staging framework. The underying semantics and staging are
two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged, only data such as sets and maps are
staged. All the monadic computation, i.e., functions passed to the monadic bind
operation @flatMap@ are statically known. This is why we can eliminate the
monadic layer and its overhead.

@@

%\todo{may be show an small example?}

@@

As the central concern of static analysis, the soundness is important for
every user who would like to use it. Here we briefly justify that how our
approach preserve the corretness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here are based on empirical evidence, a rigorous proof of soundness
perservation of the staged abstract interpreters is still an open question.

### 2019-08-15 paper/60_cases.tex
some number and typos
revision

\begin{figure*}
  \centering
  \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \begin{lstlisting}[style=extrasmall]
type CompAnalysis = Store => Store
def compProgram(prog: Expr): CompAnalysis = compCall(prog)
def compCall(call: Expr): CompAnalysis = call match {
  case Letrec(bds, body) =>
    val C1 = compCall(body); val C2 = compArgs(bds.map(_.value))
    ($\sigma$: Store) => C1(C2($\sigma$.update(bds.map(_.name), 
       bds.map(b => Set(b.value.asInstanceOf[Lam])))))
  case App(f, args) =>
    val C1 = compApp(f, args); val C2 = compArgs(args)
    ($\sigma$: Store) => C1(C2($\sigma$))
}
def compApp(f: Expr, args: List[Expr]): CompAnalysis = f match {
  case Var(x) => ($\sigma$: Store) => 
    analysisAbsApp($\sigma$.lookup(x), args, $\sigma$)
  case Op(_) => compArgs(args)
  case Lam(vars, body) =>
    val C = compCall(body)
    ($\sigma$: Store) => C($\sigma$.update(vars, args.map(primEval(_, $\sigma$))))
}
def compArgs(args: List[Expr]): CompAnalysis = args match {
  case Nil => ($\sigma$: Store) => $\sigma$
  case (arg@Lam(vars, body))::rest =>
    val C1 = compCall(body); val C2 = compArgs(rest)
    ($\sigma$: Store) => C2(C1($\sigma$))
  case _::rest => compArgs(rest)
}
  \end{lstlisting}
  \end{subfigure}
\hfill
  \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \begin{lstlisting}[style=extrasmall]
def analyzeProgram(prog: Expr, $\sigma$: Rep[Store]): Rep[Store] = 
  analyzeCall(prog, $\sigma$)
def analyzeCall(call: Expr, $\sigma$: Rep[Store]): Rep[Store] = 
  call match {
    case Letrec(bds, body) =>
      val $\sigma$_* = $\sigma$.update(bds.map(_.name), 
        bds.map(b => Set(b.value.asInstanceOf[Lam])))
      val $\sigma$_** = analyzeArgs(bds.map(_.value), $\sigma$_*)
      analyzeCall(body, $\sigma$_**)
    case App(f, args) => analyzeApp(f, args, analyzeArgs(args, $\sigma$))
  }
def analyzeApp(f: Expr, args: List[Expr], $\sigma$: Rep[Store]): 
  Rep[Store] = f match {
    case Var(x) => analyzeAbsApp(args, $\sigma$(x), $\sigma$)
    case Op(_) => analyzeArgs(args, $\sigma$)
    case Lam(vars, body) =>
      val $\sigma$_* = $\sigma$.update(vars, args.map(primEval(_, $\sigma$)))
      analyzeCall(body, $\sigma$_*)
  }
def analyzeArgs(args: List[Expr], $\sigma$: Rep[Store]): Rep[Store] = 
  args match {
    case Nil => $\sigma$
    case Lam(vars, body)::rest => 
      analyzeArgs(rest, analyzeCall(body, $\sigma$))
    case _::rest => analyzeArgs(rest, $\sigma$)
  }
  \end{lstlisting}

@@

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to contain a identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision.

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time into a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

@@

\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ... // omit code for checking caches
    _ <- put_out_cache(res._2)
    vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2)) // perform abstract garbage collection
    _ <- put_store(σ)
  } yield v
\end{lstlisting}
\vspace{-1em}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

>>>>

\begin{figure*}
  \centering
  \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \begin{lstlisting}[style=extrasmall]
type CompAnalysis = Store => Store
def compProgram(prog: Expr): CompAnalysis = compCall(prog)
def compCall(call: Expr): CompAnalysis = call match {
  case Letrec(bds, body) =>
    val C1 = compCall(body); val C2 = compArgs(bds.map(_.value))
    ($\sigma$: Store) => C1(C2($\sigma$.update(bds.map(_.name), 
       bds.map(b => Set(b.value.asInstanceOf[Lam])))))
  case App(f, args) =>
    val C1 = compApp(f, args); val C2 = compArgs(args)
    ($\sigma$: Store) => C1(C2($\sigma$))
}
def compApp(f: Expr, args: List[Expr]): CompAnalysis = f match {
  case Var(x) => ($\sigma$: Store) => 
    analyzeAbsApp(args, $\sigma$(x), $\sigma$)
  case Op(_) => compArgs(args)
  case Lam(vars, body) =>
    val C = compCall(body)
    ($\sigma$: Store) => C($\sigma$.update(vars, args.map(primEval(_, $\sigma$))))
}
def compArgs(args: List[Expr]): CompAnalysis = args match {
  case Nil => ($\sigma$: Store) => $\sigma$
  case (arg@Lam(vars, body))::rest =>
    val C1 = compCall(body); val C2 = compArgs(rest)
    ($\sigma$: Store) => C2(C1($\sigma$))
  case _::rest => compArgs(rest)
}
  \end{lstlisting}
  \end{subfigure}
\hfill
  \begin{subfigure}[h]{0.49\textwidth}
    \centering
    \begin{lstlisting}[style=extrasmall]
def analyzeProgram(prog: Expr, $\sigma$: Rep[Store]): Rep[Store] = 
  analyzeCall(prog, $\sigma$)
def analyzeCall(call: Expr, $\sigma$: Rep[Store]): Rep[Store] = 
  call match {
    case Letrec(bds, body) =>
      val $\sigma$_* = $\sigma$.update(bds.map(_.name), 
        bds.map(b => Set(b.value.asInstanceOf[Lam])))
      val $\sigma$_** = analyzeArgs(bds.map(_.value), $\sigma$_*)
      analyzeCall(body, $\sigma$_**)
    case App(f, args) => analyzeApp(f, args, analyzeArgs(args, $\sigma$))
  }
def analyzeApp(f: Expr, args: List[Expr], $\sigma$: Rep[Store]): 
  Rep[Store] = f match {
    case Var(x) => analyzeAbsApp(args, $\sigma$(x), $\sigma$)
    case Op(_) => analyzeArgs(args, $\sigma$)
    case Lam(vars, body) =>
      val $\sigma$_* = $\sigma$.update(vars, args.map(primEval(_, $\sigma$)))
      analyzeCall(body, $\sigma$_*)
  }
def analyzeArgs(args: List[Expr], $\sigma$: Rep[Store]): Rep[Store] = 
  args match {
    case Nil => $\sigma$
    case Lam(vars, body)::rest => 
      analyzeArgs(rest, analyzeCall(body, $\sigma$))
    case _::rest => analyzeArgs(rest, $\sigma$)
  }
  \end{lstlisting}

@@

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to contain a identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

@@

\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ... // omit code for checking caches
    _ <- put_out_cache(res._2)
    vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2)) // perform abstract garbage collection
    _ <- put_store(σ)
  } yield v
\end{lstlisting}
\vspace{-2em}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

### 2019-08-15 paper/80_related.tex
revision

\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation
framework for modular analysis. We borrow the notation of modular
analysis, \todo{but intentional properties are different, compare with
  summary}

@@

\paragraph{Control-flow Analysis} useful applications, why important

>>>>

\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis. We borrow the notation of modular analysis for our approach.
But the staged abstract interpreter does not generate complete summaries for
modules, instead, it is able to specialize the analysis modularly. Similar, AC
is considered as an example of this kind of mdoular analysis \cite{DBLP:conf/cc/CousotC02}.

\paragraph{Control-flow Analysis} Our case studies and evaluation focus on
control-flow anslysis for functional languages
\cite{Shivers:1991:SSC:115865.115884, Midtgaard:2012:CAF:2187671.2187672}. In
general, $k$-CFA (where $k > 0$) is considered impractical to use as a compiler
pass, due to its EXPTIME lower bound \cite{VanHorn:2008:DKC:1411204.1411243}.
However, several variants of 0CFA were developed and shown to be useful in
practice and program optimization \cite{Adams:2011:FTR:2048066.2048105,
Bergstrom:2014:PEH:2628136.2628153, ashley:practical, Reppy:2006:TCA:1159876.1159888}.

@@

### 2019-08-15 paper/90_conc.tex
revision

In this paper, we first present the implementaion of first Futamura projection
on abstract interpreters. To realize it, we design a generic interpreter that
abstracts over both binding-time and value domains; by instantiating it
differently, the staged abstract interpreter can be readily derived. We also
conduct case studies and empirical evaluation, demonstrating a wide
applicability of our approach and evaluating the performance improved by staging.

>>>>

In this paper, we present a simple and elegant framework that extends the first
Futamura projection to abstract interpreters.
To realize it, we design a generic interpreter that abstracts over both
binding-time and value domains; by instantiating it with staged monads for
abstract interpretation, the staged abstract interpreter can be readily derived.
The meta-programming approach brings new insight and implementation techniques
to optimize static analyzers. The optimization can be realized in a completely
orthogonal way to the interpreter and abstract semantics.
In our empirical evaluation, staging improves an order of magnitude of
performance in terms of running time.
